---
title: "C03_assignment.Rmd"
output: github_document
---
#set up r and name species 

```{r setup, include=FALSE}
source("setup.R")
SPECIES2 = "Doryteuthis pealeii"
```

## Read species 

```{r read_obs}
obs = read_obis(SPECIES2)
obs
```
This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Summary of observations 

You can include R code in the document as follows:

```{r cars}
summary(obs)
```
#filter out na values in dataset
```{r obs_filter_date}
obs = obs |>
  filter(!is.na(eventDate))
summary(obs)
```


## plot of observation 
```{r plot_year, echo=FALSE}
ggplot(data = obs,
       mapping = aes(x = year)) + 
  geom_bar() + 
  geom_vline(xintercept = c(1982, 2013), linetype = "dashed") + 
  labs(title = "Counts per year")
```
#load data 
```{r load_obs_mask}
coast = read_coastline()
obs = read_observations(scientificname = "Doryteuthis pealeii")
db = brickman_database() |>
  filter(scenario == "STATIC", var == "mask")
mask = read_brickman(db)
```

#change coordintes of observation 
```{r all_observations_count}
LON0 = -67
LAT0 = 46
all_counts = count(st_drop_geometry(obs), month)
all_counts
```

#GG plot created for species observation presence 
```{r_ggplot}
ggplot() +
  geom_sf(data = obs, alpha = 0.2, shape = "circle small", size = 1) +
  geom_sf(data = coast, col = "green") +
  geom_text(data = all_counts,
            mapping = aes(x = LON0, 
                          y = LAT0, 
                          label = sprintf("n: %i", .data$n)),
                          size = 3) + 
  labs(x = "Longitude", y = "Latitude", title = "All observations") +
  facet_wrap(~month)
```
#thin observations filtering out months without data (should be none)
```{r thin_observations}
thinned_obs = sapply(month.abb,
               function(mon){ 
                 thin_by_cell(obs |> filter(month == mon), mask)
               }, simplify = FALSE) |>
  dplyr::bind_rows() 
thinned_counts = count(st_drop_geometry(thinned_obs), month)

ggplot() +
  geom_sf(data = thinned_obs, 
          alpha = 0.2, 
          shape = "circle small", 
          size = 1) +
  geom_sf(data = coast, col = "orange") +
  geom_text(data = thinned_counts,
            mapping = aes(x = LON0, 
                          y = LAT0, 
                          label = sprintf("n: %i", .data$n)),
                          size = 3) + 
  labs(x = "Longitude", y = "Latitude", title = "Thinned observations") +
  facet_wrap(~month)
```
#ggplot bias map 
```{r bias_map}
bias_map = rasterize_point_density(obs, mask) 

ggplot() +
  geom_stars(data = bias_map, aes(fill = count)) +
  scale_fill_viridis_b(na.value = "transparent") +
  geom_sf(data = coast, col = "orange") + 
  labs(x = "Longitude", y = "Latitude", title = "Bias map using all observations")
```
#explore background 
```{r explore_background}
nback_avg = mean(all_counts$n) |>
  round()
nback_avg
```
#sample of background 
```{r sample_background}
obsbkg = sapply(month.abb,
    function(mon){ 
      sample_background(thinned_obs |> filter(month == mon), # <- just this month
                       bias_map,
                       method = "bias",  # <-- it needs to know it's a bias map
                       return_pres = TRUE, # <-- give me the obs back, too
                       n = nback_avg) |>   # <-- how many points
        mutate(month = mon, .before = 1)
    }, simplify = FALSE) |>
  bind_rows() |>
  mutate(month = factor(month, levels = month.abb))
obsbkg 
```
#check tallying for each month
```{r nback_avg}
count(st_drop_geometry(obsbkg), month, class)
```

#ggplot background 
```{r plot_presence_background}
ggplot() +
  geom_sf(data = obsbkg, 
          mapping = aes(col = class),
          alpha =  0.4, shape = "circle small", size = 1) +
  geom_sf(data = coast, col = "orange")  + 
  labs(x = "Longitude", y = "Latitude", title = "All") +   
  theme_bw() + 
  scale_fill_okabe_ito() +  
  facet_wrap(~month)

```

#creating model for species 
```{r save_model_input}
write_model_input(obsbkg, scientificname = "Doryteuthis pealeii")
```
#save model 
```{r read_model_input}
x = read_model_input(scientificname = "Doryteuthis pealeii")
```
#generate variables in brickman database
```{r read_brickman}
db = brickman_database() |>
  dplyr::filter(scenario == "PRESENT", interval == "mon")
present = read_brickman(db)
```
#generate pairs plot for variables in brickman database vs. observations 
```{r pairs}
pairs(present)
```
#filter significant correlation values 
```{r filter_collinear}
keep = filter_collinear(present, method = "cor_caret", cutoff = 0.65)
keep
```









